---
title: 'Project 01: MLB Home Runs'
author: "Brooke Coneeny"
date: "1/10/2021"
output: 
  pdf_document:
    highlight: pygments
    dev: cairo_pdf
fontsize: 11pt
geometry: margin = 1in
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Overview:

For this project I chose to replicate and re-design a visualization which shows how the number of MLB home runs per year has changed over time.


Link to the article containing the graph and data: https://cnsmaryland.org/2019/10/01/a-look-into-major-league-baseballs-massive-home-run-surge-in-2019/

The original data set consists of 2 variables (Year & Number of HRs) and 149 observations (1871-2020). I created a new variable, Era, which categorized what era of baseball each of the years fell into. I used this new data set throughout the project, but the effects of the change can only be seen in the glow-up graph. 

```{r homeruns}
#Load in data
library(readxl)
homeruns <- read_excel("HOMERUNS.xlsx")
```

**This is the code for the replica I created:**
```{r replica}
#Create Replica Plot
library(tidyverse)
library(directlabels)
ggplot(data = homeruns, mapping = aes(x = Year, y = HR)) +
  geom_line(color = "skyblue", size = 1) +
  scale_x_continuous(breaks = seq(1880, 2020, 10)) +
  scale_y_continuous(breaks = seq(0,6500,500)) +
  labs(title = "MLB Home Runs by Year", size = 5) +
  theme_classic() +
  theme(
    panel.grid.major.x = element_blank(), 
    panel.grid.major.y = element_line(size = .1, color = "lightgrey"),
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    axis.ticks = element_blank(),
    axis.line.y = element_blank(),
  ) +
  annotate("text", x = 2020, y = 7000, label = "HR")
```
  

**This is the code for the glow-down I created:** 
```{r glow-down}
#create Glow-Down plot
ggplot(data = homeruns, mapping = aes(x = Year, y = HR)) +
  geom_point(color = "blue4") +
  scale_x_continuous(breaks = seq(1880, 2020, 10)) +
  scale_y_continuous(breaks = seq(0,6500,500)) +
  labs(title = "MLB Home Runs by Year", size = 5) +
  theme_classic() +
  theme(
    panel.grid.major.x = element_blank(), 
    panel.grid.major.y = element_blank(),
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    axis.ticks = element_blank(),
    axis.line.y = element_blank(),
    panel.background = element_rect(fill = "gray35"),
    plot.background = element_rect(fill = "royalblue2"),
    text = element_text(family = "Impact")
  )
```


**This is the code for the glow-up I created:**
```{r glow-up}
#create Glow-Up plot
library(extrafont)
ggplot(data = homeruns) +
  geom_col(mapping = aes(x = Year, y = HR, fill = Era)) +
  scale_fill_brewer(palette = "Paired") +
  geom_line(mapping = aes(x = Year, y = HR), color = "black") +
  scale_x_continuous(breaks = seq(1880, 2020, 10)) +
  scale_y_continuous(breaks = seq(0,6500,500)) +
  labs(
    title = "MLB Home Runs by Year", 
    subtitle = "Total number of combined home runs by all teams in the MLB, categorized by Baseball-Era"
    ) +
  theme(
    plot.background = element_rect(fill = "gainsboro"),
    panel.background = element_rect(fill = "gainsboro"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    text = element_text(family = "Century", size = 10),
    plot.title = element_text(size = 20),
    legend.background = element_rect(fill = "gainsboro")
  )
```


